manuscript_v1

### 2022-01-01 manuscript_v1.Rmd
cut text round 1: methods

In general, interspecific competition is likely to be higher among closely related species due to similar resource use patterns which would lead to smaller overall habitat breadths and overlaps, whereas without competitive effects related species can be expected to overlap more since they have similar basic feeding patterns. Alatalo (1981) found evidence for the prevalence of interspecific competition in the seasonal environment of Finland and noted that this ecological process might be working at two different time scales: either in current ecological time as exemplified by interspecific territoriality in some bird species (e.g., Orians and Willson, 1964) or during the evolution of feeding adaptations of the species—for example, tit species are better adapted to exploit resources in their preferred habitats than in other habitats (Partridge, 1976; Snow, 1954).

@@

The study was conducted over the course of three months (13 weeks), from the first week of July 2020 to the first week of October 2020, in Branišovský les (les: ‘forest’), a typical structurally manipulated and commercially exploited patch of Central European temperate forest situated in České Budějovice, a city in the South Bohemian region (Jihočeský kraj) of the Czech Republic. The forest is located on the western edge of the city between the housing estate Máj and the villages of Branišov, Mokré and Třebín. Together with a southern section called Homolský les, it forms a large forest stretch of around 800 ha segregated from other nearby forests by extensive farmland, large fishponds and settlements (as well as the small airport to the south of Homolský les). Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is highly heterogeneous owing to the long history of forestry regimes as well as its connection with rural and military Czech history (having housed, i.a., a military garrison, and functioned as a military shooting range from the times of the Austro-Hungarian empire), and ranges from pure spruce plantations to mixed stands dominated by oak.

As such, the major tree species include conifers like Norway spruce Picea abies, Scots pine Pinus sylvestris, European larch Larix decidua and silver fir Abies alba, and broad-leaved species such as pedunculate/European oak Quercus robur, sessile/Cornish oak Q. petraea, northern red oak Q. rubra, silver birch Betula pendula, quaking/Eurasian aspen Populus tremula, common/European beech Fagus sylvatica, rowan or European mountain ash Sorbus aucuparia, small-leaved lime or littleleaf linden Tilia cordata, common/European black alder Alnus glutinosa, Norway maple Acer platanoides and sycamore maple A. pseudoplatanus. The commonly found shrub species are alder buckthorn Frangula alnus, rowan Sorbus aucuparia and black/European elderberry Sambucus nigra, while ground layer vegetation is restricted to graminoids like sedges Carex spp. (mostly C. brizoides), bramble Rubus sp. and European blueberry Vaccinium myrtillus.

## Study design

The entire forest stretch—with the exception of parts of Homolský les in the south, some of which were either too disturbed or had restricted access (being near the airport)—was effectively covered by selecting 32 evenly distributed circular sampling points of 30 m radius (Figure 1). The total forest area under consideration was thus around 400 ha, while the absolute cumulative area covered by my sampling points was approximately 9 ha, as they were chosen such that each point was at least 250 m away from its neighbours (although this boundary was lowered to 200 m in three cases). Clear cuts were avoided as much as possible. Therefore, the selection of sampling points was not done at random.

Figure 1: Map showing 32 sampling points evenly distributed over the study area: Branišovský les, Czech Republic.

Before the beginning of the main study, each point was visited and preparations were made. GPS coordinates were recorded from the centre of each point using the GPS Essentials Android application and saved for reference. The centre of a point was signified using red-and-white tape tied and/or stuck to the trunk or branch of a tree, which ensured identifiability and hence constancy. A radius of 20 m from the centre was also denoted using tape stuck/tied to trees in two opposite directions. In addition, three equidistant sub-points within 10–15 m of the centre were chosen to sample canopy cover from.

To measure canopy cover, first, photographs of the canopy were taken from each of the three sub-points (named A, B and C) using a Nikon D3200 camera with a standard 18–55 mm kit lens set to the 18 mm focal length. Measurements were always taken by the same observer (myself) from the same locations (sub-points identified by the tapes) and at the same height (around 1.9 m above the ground; assured by holding the camera horizontally tethered to my neck by the strap, and positioning both the camera and my person vertically upright and/or perpendicular to the ground, whichever appropriate to the topography). The photographs were then analysed using ImageJ software (Rasband, 1997) by converting to binary black-and-white images (see Forester, 2011) and calculating the percentage of cover across the whole image (see Figure S1). The measurements from A, B and C sub-points were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity).

Aside from canopy cover, other vegetation characteristics were also recorded initially. The tree, shrub and herb species were identified, and the vertical strata to which they belonged were noted. Some species like S. aucuparia are highly variable and commonly occur as both shrubs and trees. Since my primary question concerned the structural characteristics of vegetation rather than their taxonomic or biological traits, vertical stratification was given precedence over the tree-shrub-herb classification, and the various plants in each of my sampling points were categorised as follows. Plants of height ≤ 1 m above the ground belonged to the ‘Herb’ layer (this therefore included both woody- [like Rubus] and non-woody stemmed [like graminoids] plants, and did not exclude graminoid leaves that extended beyond 1 m from the ground); plants 1–4 m in height belonged to the ‘Understorey’ (this therefore included shrubs like S. aucuparia and F. alnus as well as young individuals of trees like P. abies); and plants ≥ 4 m in height belonged to the ‘Canopy’ (this therefore occasionally included S. aucuparia).

The species information was used to calculate plant species richness and diversity in each of the points. Shannon’s diversity index (Shannon, 1948) was used because it emphasises the richness component of diversity, and is more sensitive to rare types and less to dominant ones, and is therefore generally recommended for ecological questions like the ones in this present study (see Nagendra, 2002). The number of ‘Canopy’ and ‘Understorey’ plants within the 20 m radius was counted and used to calculate density per 100 m2. Counts were not taken for the ‘Herb’ layer due to the difficulty in some cases in determining what constitutes a single stem. Moreover, it is unlikely that birds track ground layer vegetation at the level of individual stems, and so this was not pertinent to my primary question. Instead, the species that covered >50% of the ground was considered to be the ‘Dominant Ground Vegetation’ (in some cases, this was moss) and this was used in analyses. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality of and disturbance within the forest patch (see e.g., Berdugo et al., 2018; Cornelissen et al., 2007).

The proportion of ‘Canopy’ trees that are deciduous was calculated as this has been shown to be more important than measures of tree diversity (Berg, 1997). Finally, the habitat class of each of the 32 points was recorded as ‘Interior’, ‘Road’ or ‘Edge’ according to whether or not asphalt roads and forest-to-farmland transitions occurred within 150 m from the centre of the point. This was done using an aerial map (Mapy.cz) in which asphalt roads, farmlands and forests are all clearly visible and distinguishable.

Care was taken to ensure that the selected set of points represented sufficiently well the natural range and diversity of habitat characteristics in the forest (see Figure S2 for some comparisons). This was confirmed by an NMDS plot of the sampling points with their measured characteristics; there was no major clumping of points and no observable residual pattern in the ordination, suggesting that the points and their features were sufficiently diverse. Linear correlations of the measured habitat variables were tested using Pearson’s product moment correlation before building the statistical models, so as to spot and exclude some of the correlated variables and thereby avoid multicollinearity. For instance, the mean canopy cover was found to correlate significantly with tree density and with canopy heterogeneity, but tree density and canopy heterogeneity in turn were not correlated (Figure S6).

I recorded canopy cover for all my sampling points every week throughout my study period, as I had originally expected to capture a decline in cover due to autumn senescence. However, most of the deciduous trees in my points still bore leaves even during my last sampling (in the first week of October) and the weekly measurements did not show any discernible negative trend. At the same time, there was variation in each of the unique values (A, B and C) around a mean, likely due to varied weather conditions in each of the samplings affecting the positions of leaves (e.g., winds) and lighting. Hence, I decided to average the 13 weekly measurements of both mean canopy cover and canopy heterogeneity for each sampling point to provide more accurate measures of both. Furthermore, along with the bird counts and canopy cover measurements, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns along the season. The presence of intense shrub herbivory in some cases was recorded. The demarcating tapes were monitored and replaced if required.

@@

To answer the primary question regarding habitat selection patterns of post-breeding forest birds, I used point counts to observe birds at each of the 32 sampling points every week for 13 weeks (8 July 2020–1 October 2020). Each point being located at least 250 m from its neighbours assured spatiotemporal independence of consecutive point counts. Sampling was done on 2–3 days in the middle of the week while the remaining days served to delineate and separate two successive ‘temporal units’. This enabled meaningful comparisons of bird abundances across weeks, as each would be sufficiently independent of the other and would thus represent a distinct time point (Table 1).

To ensure logistical feasibility of the sampling design, bird counts were conducted by two observers, myself and Inga Freiberga. The binoculars used were Nikon Aculon A211 8x42 and Nikon Prostaff 7s 8x42 respectively. Inga has a Master’s degree in Ornithology and years of active experience with birdwatching and mist-netting across Europe. I had spent 170+ hours in the field birdwatching by the start of the study and additionally, bird songs and calls were regularly reviewed using online repositories such as Macaulay Library and xeno-canto, in order to familiarise myself with the seasonally changing vocalisations and thus ensure best possible bird detection.

We, myself and Inga, each conducted 6–10 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Overcast and rainy days were avoided. The subset of points that each observer sampled was alternated every week, and the order in which the points were sampled was reversed as well. The counts were conducted with the observer standing in the centre of the point, and each count was of 20 minutes, chosen after a few recce counts as the optimal count duration for this study period. Count-specific information like observer name, date, week, weather (overcast, cloudy, partially cloudy or clear sky), wind (strong winds, light winds or no winds) and visibility (misty or clear), and start and end times were noted in each count. The 20-minute duration was divided into four intervals of five minutes each. All birds heard or seen were recorded, along with complementary information associated with each detection such as the time interval (one of the four 5-minute units), distance class (see next), type of detection (visual or auditory or both) and behaviour such as flyover (over the forest, within the forest canopy or not flying) and foraging (yes or no). If foraging, the plant species on which the bird was found foraging was noted as well as whether or not it was feeding on fruits.

Table 1: The continuous measure of time (Week) used to test for patterns in habitat selection, with corresponding weeks (starting on Monday) in the Gregorian calendar.

Week	Corresponding Gregorian week
1	6 July 2020–12 July 2020
2	13 July 2020–19 July 2020
3	20 July 2020–26 July 2020
4	27 July 2020–2 August 2020
5	3 August 2020–9 August 2020
6	10 August 2020–16 August 2020
7	17 August 2020–23 August 2020
8	24 August 2020–30 August 2020
9	31 August 2020–6 September 2020
10	7 September 2020–13 September 2020
11	14 September 2020–20 September 2020
12	21 September 2020–27 September 2020
13	28 September 2020–4 October 2020

Distance classes referred to five radial bins from the centre of the point: 0–10 m, 10–20 m, 20–30 m, 30–50 m and >50 m. This was done in case detection probabilities needed to be calculated, so additionally, birds were observed for whether or not they entered the 30 m radius (i.e., entered the point) at any time during the count, as this would signify the bird’s ‘use’ of the habitat. The distance sampling approach using detection probabilities was not appropriate for the study due mainly to insufficient detections. Moreover, estimating true population sizes of birds was not an objective of the study, and hence raw abundances from the point counts were still suitable. Since a large majority of the detections were auditory, errors in detection were ascribed more to differences in bird vocalisation behaviour with the progressing season than to differences in vegetation characteristics of sampling points. Hence, since analyses were done at the scale of weeks, inferences are not expected to be highly unreasonable.

Distances beyond 20 m from the centre, particularly the third class of 20–30 m, were estimated with the help of the red-and-white marking tapes as references—the observer would mentally note the distance from the centre of the point to one of the tapes demarcating the point boundary, halve it, and then transplant this half to the other side of the tape so as to estimate the 30 m mark. The accuracy of such distance estimation obviously decreases with increasing distance, hence the distance bins beyond 30 m were wider. All birds were identified to species level when possible, but some closely related cryptics such as Phylloscopus spp., Certhia spp., Regulus spp. and Dendrocopos spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as ‘spuhs’. Only those birds that entered the 30 m radius during point counts and were not flying over the forest have been considered for analysis in the present study.

@@

## Predation

Survey of predation pressure was conducted at the same 31 points as the ornithological surveys (one point excluded due to the absence of suitable foliage). Siddharth Khopkar assisted with this experimental part, while Kateřina Sam confirmed the accuracy of bite identifications. We used dummy caterpillars exposed on the trees and shrubs to monitor attacks by natural enemies. Caterpillars were made from natural-looking, dark green-coloured modelling clay (Koh-i-Noor Hardtmuth, České Budějovice, Czech Republic), which is malleable, oil-based and non-toxic. We modelled artificial caterpillars by pressing the plasticine through a stainless-steel clay extruder gun (Antoble) to ensure that each caterpillar had an absolutely smooth surface. Artificial caterpillars were 20 mm long and 3 mm in diameter. The artificial caterpillars resembled generic non-conspicuous lepidopteran larvae commonly found on oaks in Central Europe (e.g., autumnal moth Epirrita autumnata, winter moth Operophtera brumata). This method has been successfully used in previous studies (e.g., Howe et al., 2009; Drozdova et al., 2013; Mrazova and Sam, 2018; Sam et al., 2015).

The survey started with a one-month delay respective to the ornithological part of the study, and each sampling usually took two days to finish. Thus, the survey spanned from Week 5 to Week 12 of the main study. During experimental days 1 and 2 (27th and 28th of July 2020), we exposed caterpillars on shrubs and trees and other foliage, within the 30 m radius of the centre of the point, at heights between 1 and 2.5 m above the ground. We pinned the 20 dummy caterpillars on bases of leaf stalks, at a minimal distance of 1 m from each other.

A week later, we visited all the points again, checked for missing and damaged dummy caterpillars, and collected them for further reference. Caterpillars without visible damage were left unchanged. Damaged ones were replaced by new ones, placed within a 50 cm radius but not on the same leaf. An artificial caterpillar was considered as damaged by insectivorous birds if it had some unequivocal attack marks distinctive to birds (Low et al., 2014). Caterpillars attacked by arthropods only were considered to be attack-free. Some dummy caterpillars were missing, and as we were not able to identify their fates and times of disappearance, they were excluded from the dataset prior to calculations of the predation rate in proportions (i.e., missing caterpillars were considered as having never been exposed). At any given time, there was a total of 620 artificial caterpillars exposed to predators. Given that dummy caterpillars were exposed eight times by the end of the study period, with a total exposition time of 66 days, this yielded 4,960 potential predation events and 40,920 caterpillar exposition days respectively.

## Analysis

The bird data containing all observations from the point counts was filtered to exclude the birds that did not enter the 30 m radius during the count, those that flew over the forest canopy and those that were not identified to genus level. To assess patterns of habitat selection by birds, analyses were conducted at three main levels of response, using generalised linear mixed models (GLMM). Mixed models were used to account for the repeated samplings, attributing a random variability to them so as to avoid pseudoreplication. Bird abundance at various levels was used as the response variable. The measured habitat variables without the ones excluded due to correlations represented the pool of candidate predictors. This set of plausible predictors was then used to build models using an information-theoretic approach (Burnham and Anderson, 1998), viz., the Akaike Information Criterion (Akaike, 1973, 1974). More specifically, the corrected AIC (AICC) was used (Hurvich and Tsai, 1989). Traditional parametric tests and p-values were avoided as they are not reliable for complex models such as GLMMs. Non-focal variables were tested for effects first, after which the focal habitat characteristics were tested.

To assess differences in bird predation rates across the study area and with time, analysis was done using GLMMs with the habitat variables as predictors. Additionally, correlation between bird predation rates and bird abundances were checked using Pearson’s product moment correlation. All analyses were done using R version 4.0.2 (R Core Team, 2020). Individual figures with multiple graphs were composed using the patchwork package for R (Pedersen, 2020).

## Regressions of bird abundances with habitat variables

First, I investigated whether there were any patterns of habitat selection at the highest level of all bird species by summing all the individual species abundances per point count to obtain one abundance value for each count. This ‘bird abundance’ was then used as the response variable in the models. GLMMs were constructed using the lme4 package (Bates et al., 2015), starting with a null model and building up from it. The null model included observer as a fixed effect and sampling points and days as crossed random effects. The residuals were modelled as coming from a Poisson distribution with a log link function. There was a prominent observer effect, with me making more detections on average than Inga (see Figure S4). Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day was used as a random effect to account for the random variability in each unique day.

Next, the week variable, which clearly had a strong effect (abundances were higher towards the end of the study period; noticed even during the bird counts), followed by other non-focal (‘nuisance’) variables such as weather, visibility and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in) were added to the null models and comparisons were made of AICC values of all the models using the AICctab() function from the bbmle package (Bolker et al., 2020). If the addition of a variable did not result in a substantial decrease of AICC respective to the number of degrees of freedom it used (following the rule of thumb of 2 dAICC per df), it was discarded, and otherwise it was retained in the model. Such a stepwise procedure was repeated until the final model was obtained. However, the null model in the beginning was not compared with a simpler model, as the variables were all assumed to have an effect and discarding them would only be statistically incorrect.

None of the non-focal variables ended up in the model, and finally other focal variables were tested including habitat class, presence of moss, log-transformed variables of canopy heterogeneity and tree density and square root-transformed understorey density (transformations done to prevent outliers from gaining too much leverage), scaled proportion of deciduous trees, tree species diversity and richness, and dominant herb. Understorey richness and diversity were not tested because the sampling points did not have sufficient diversity in understorey species and moreover, it was unlikely that birds would track, for instance, spruce saplings (which indeed belonged to my ‘Understorey’ classification). Instead, understorey density was thought to be a meaningful and more important predictor. For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. If a variable was found to be important, its interaction with ‘Week’ was next tested, this being one of the primary questions of my study (i.e., whether or not certain habitat selection criteria changed with the progress of the season).

Model diagnostics of all models were explored using the DHARMa package (Hartig, 2020) for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. Since the model with the default optimiser was facing convergence problems, the BOBYQA optimiser was used which resolved the issue. Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was obtained.

The next level of response investigated was the guild level. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Models were then built separately for the two guilds that had the highest number of overall detections, insectivores (rather, invertebrate-feeders) and omnivores. Using the lme4 package resulted in several errors and warnings regarding convergence and singular fits, but the same model specifications using the glmmTMB package (Brooks et al., 2017) did not, and hence the latter was used in further steps. The data summarised at guild level was insufficient for the proper utilisation of two crossed random effects (as was done in the first model), so the random effect of day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates so was not considered to be an issue. Model building and checking of model diagnostics was done in the same way as in the first case.

To explore whether patterns differed between migrant and non-migrant species, the bird data was summarised by migration strategy and attempts were made to build GLMMs using the glmmTMB package. However, while a good fitting model was obtained for non-migrants (which had the highest number of overall detections), data for the other two categories seemed to be unsuitable and insufficient. The data showed significant one-inflation when tested (using functions in the DHARMa package) relative to both Poisson and negative binomial distributions, i.e., the number of ‘1’ count observations was too high for a standard Poisson or a standard negative binomial. This one-inflation in turn also caused underdispersion, which was removed neither on switching to distributions suited to underdispersed data (such as the Generalised Poisson and Conway-Maxwell Poisson), nor on introducing a zero-inflation term. The same problem occurred during attempts to build models at the species-level, i.e., looking at patterns of habitat selection for certain selected species (incidentally, the ones with most overall detections). The results of these questions were expected to be interesting, but as the data didn’t seem much suited to a binomial distribution either, I accepted that better data would be required and thus this approach was abandoned. Finally, separate models were built to test whether bird abundances were linked to the phenology of the fruiting plants Rubus sp., Vaccinium sp., F. alnus and S. aucuparia but there was no detectable relationship.

## Ordinations of selected bird species communities

In order to gain a qualitative idea of habitat selection at the species level, communities defined by the preselected species—four invertebrate-feeders (European robin Erithacus rubecula, treecreepers Certhia spp., common chiffchaff/willow warbler Phylloscopus spp., and goldcrest/common firecrest Regulus spp.) and four omnivores (European great tit Parus major, Eurasian blue tit Cyanistes caeruleus, Eurasian nuthatch Sitta europaea, and great-/middle spotted woodpeckers Dendrocopos spp.)—across the entire study area were compared between the start and the end of the study period. Data from the first three weeks of the study (1–3) were aggregated to represent the ‘post-breeding period’ (PB) and data from the last three (11–13) were aggregated to represent the ‘pre-migratory period’ (PM). Abundances in the same points in different weeks were summed. Principal Component Analyses (PCAs) scaled by species were done using the vegan package (Oksanen et al., 2020) and the ordinations were plotted using the first two principal components. The measured habitat variables were fitted to the ordination diagrams using the envfit() function from vegan, and the two plots were then compared to see similarities and differences in the bird species between the two periods and how occurrences of these species over the study area were generally structured.

## Bird predation across the points and over time

First, the variation in bird predation rates across the sampling points and over the course of the 13 weeks of the study was explored using a generalised linear model (GLM) with binomial error distribution. A binomial response variable of total bird predations (number of caterpillars attacked by birds) over total non-predations was modelled against fixed effects of point and week. This model was updated to use a second-order polynomial of the week, which provided an even better fit (11 dAICC on 1 df).

However, since it was now ascertained that the points were indeed different in terms of bird predation, it was decided to replace the point variable with habitat variables to create a more parsimonious as well as informative model, while accounting for random variation by using point as a random effect. Variables from the same pool as before were tested—expecting plant diversity to be an important predictor of predation rates (e.g., Leles et al., 2017)—in generalised linear mixed models with binomial error distribution using the lme4 package. Finally, Pearson’s product moment correlation between bird abundances and the predation rates in each point in each week was calculated.

@@

There were 494 records of birds foraging, although just 2 of them were of birds feeding on fruits. Even in the full unrestricted dataset, there are records of just 14 birds feeding on fruits, and half of these birds were from a single observation. The fruiting plant species most birds fed on was rowan S. aucuparia although one record was of a blackbird Turdus merula feeding on Rubus sp. Most of the birds feeding on fruits were thrushes, but one instance was of a Eurasian blackcap Sylvia atricapilla and another of two Eurasian jays Garrulus glandiarus. Although this data is very limited and cannot be used in further analyses, the tiny proportion of overall foraging observations that were of frugivory is certainly an interesting point.

>>>>

In general, interspecific competition is likely to be higher among closely related species due to similar resource use patterns which would lead to smaller overall habitat breadths and overlaps, whereas without competitive effects related species can be expected to overlap more since they have similar basic feeding patterns. Alatalo (1981) found evidence for the prevalence of interspecific competition in the seasonal environment of Finland and noted that this ecological process might be working at two different time scales: either in current ecological time as exemplified by interspecific territoriality in some bird species (e.g., Orians and Willson, 1964) or during the evolution of feeding adaptations of the species---for example, tit species are better adapted to exploit resources in their preferred habitats than in other habitats (Partridge, 1976; Snow, 1954).

@@

The study was conducted over the course of three months (13 weeks), from the first week of July 2020 to the first week of October 2020, in a typical structurally manipulated and commercially exploited patch of Central European temperate forest in České Budějovice, Czech Republic. The forest is located on the western edge of the city, and represents a semi-urban forest stretch of around 800 ha segregated from neighbouring forests by extensive farmland, large fishponds and settlements. Situated at an altitude of approximately 400 m asl, it is technically a mixed forest, but is highly heterogeneous owing to the long history of forestry regimes as well as its connection with rural and military Czech history, and ranges from pure spruce plantations to mixed stands dominated by oak.

As such, the major tree species include conifers like Norway spruce *Picea abies*, Scots pine *Pinus sylvestris*, European larch *Larix decidua* and silver fir *Abies alba*, and broad-leaved species such as pedunculate/European oak *Quercus robur*, sessile/Cornish oak *Q. petraea*, northern red oak *Q. rubra*, silver birch *Betula pendula*, quaking/Eurasian aspen *Populus tremula*, common/European beech *Fagus sylvatica*, rowan or European mountain ash *Sorbus aucuparia*, small-leaved lime or littleleaf linden *Tilia cordata*, common/European black alder *Alnus glutinosa*, Norway maple *Acer platanoides* and sycamore maple *A. pseudoplatanus*. The commonly found shrub species are alder buckthorn *Frangula alnus*, rowan *Sorbus aucuparia* and black/European elderberry *Sambucus nigra*, while ground layer vegetation is restricted to graminoids like sedges *Carex* spp. (mostly *C. brizoides*), bramble *Rubus* sp. and European blueberry *Vaccinium myrtillus*.

The entire forest stretch---with the exception of some parts either too disturbed or having restricted access---was effectively covered by selecting 32 non-random, evenly distributed circular sampling points of 30 m radius (Figure 1). The total forest area under consideration was thus around 400 ha, while the absolute cumulative area covered by my sampling points was approximately 9 ha, as they were chosen such that each point was at least 250 m away from its neighbours (although this boundary was lowered to 200 m in three cases). Clear cuts were avoided as much as possible. Care was taken to ensure that the selected set of points represented sufficiently well the natural range and diversity of habitat characteristics in the forest.

![Map showing 32 sampling points evenly distributed over the study area: Branišovský les, Czech Republic.](map_studyarea.png)

## Vegetation surveys

Vegetation features thought to be relevant to birds in the post-breeding period were chosen in order to characterise each of the various habitat patches. The tree, shrub and herb species were identified, and the vertical strata to which they belonged were noted. Regardless of taxonomic and functional identity, plants of height ***≤*** 1 m above the ground belonged to the ‘Herb’ layer (this therefore included both woody- [like *Rubus*] and non-woody stemmed [like graminoids] plants, and did not exclude graminoid leaves that extended beyond 1 m from the ground); plants 1--4 m in height belonged to the ‘Understorey’ (this included shrubs like *S. aucuparia* and *F. alnus* as well as young individuals of trees like *P. abies*); and plants ***≥*** 4 m in height belonged to the ‘Canopy’ (this occasionally included *S. aucuparia*).

The species information was also used to calculate plant species richness and diversity (Shannon’s diversity index; Shannon, 1948) in each of the points. Counts of ‘Canopy’ and ‘Understorey’ plants were taken and used to calculate density per 100 ***m2***. Counts were not taken for the ‘Herb’ layer due to the difficulty in some cases in determining what constitutes a single stem and since it is unlikely that birds track ground layer vegetation at the level of individual stems. Instead, the species that covered >50% of the ground was considered to be the ‘Dominant Ground Vegetation’ (in some cases, this was moss) and this variable was also used in analyses. In addition, presence or absence of moss in the point was noted separately regardless of whether or not it was the dominant ground layer vegetation, as its presence usually indicates the quality of and disturbance within the forest patch (see e.g., Berdugo et al., 2018; Cornelissen et al., 2007). The proportion deciduous tress in the 'Canopy' was calculated as this has been shown to be a more important factor than measures of tree diversity (Berg, 1997). Furthermore, the habitat class of each of the 32 points was recorded as ‘Interior’, ‘Road’ or ‘Edge’ according to whether or not asphalt roads and forest-to-farmland transitions occurred within 150 m from the centre of the point. This was done using an aerial map (Mapy.cz) in which asphalt roads, farmlands and forests are all clearly visible and distinguishable.

Canopy cover at a sampling point was measured using canopy photographs taken at three equidistant sub-points within 10–15 m of the centre, which were then analysed using ImageJ software (Rasband, 1997) by converting to binary black-and-white images (see Forester, 2011) and calculating the percentage of cover across the whole image (see Figure S1). The measurements from each sub-point were averaged to represent mean canopy cover of a point, while the standard deviation of the three values represented the heterogeneity in canopy cover in the point (high values signifying high heterogeneity). Such measurements were taken once in each sampling week. Since most deciduous trees bore leaves even in the last week of sampling, the 13 weekly measurements of both mean canopy cover and canopy heterogeneity for each sampling point were averaged to provide more accurate measures of both. Lastly, phenological characteristics of leaves, flowers and fruits of the plants in each point were also recorded weekly on an ordinal scale in order to track fruiting, flowering and leaf senescence patterns across the study period.

***Table 1: The continuous measure of time (Week) used to test for patterns in habitat selection, with corresponding weeks (starting on Monday) in the Gregorian calendar.***

@@

We surveyed bird abundance by means of fixed-radius point counts at each of the 32 sampling points every week for 13 weeks (8 July 2020--1 October 2020). Each point being located at least 250 m from its neighbours assured spatiotemporal independence of consecutive point counts. Sampling was done on 2--3 days in the middle of the week while the remaining days served to delineate and separate two successive ‘temporal units’. This enabled meaningful comparisons of bird abundances across weeks, as each would be sufficiently independent of the other and would thus represent a distinct time point (Table 1).

We conducted 12--20 bird counts per sampling day, during the first five hours after sunrise (which shifted forward by around 10 minutes every week). Overcast and rainy days were avoided. We recorded the presence and counts of all birds heard or seen at each point, during a period of 20 minutes divided into four intervals of five minutes each. Additionally, complementary information for each detection, like type of detection (visual, auditory or both), behaviour such as flyover (over the forest, within the forest canopy or not flying) or foraging (yes or no), and distance class were noted. Distance classes referred to five radial bins from the centre of the point: 0--10 m, 10--20 m, 20--30 m, 30--50 m and >50 m. Additionally, birds were observed for whether or not they entered the 30 m radius (i.e., entered the point) at any time during the count, as this would signify the bird’s ‘use’ of the habitat. All birds were identified to species level when possible, but some closely related cryptics such as Phylloscopus spp., Certhia spp., Regulus spp. and Dendrocopos spp. are easy to confuse and difficult to identify correctly with just the occasional post-breeding vocalisations, and hence were often recorded as ‘spuhs’.

@@

## Data analysis

To assess patterns of habitat selection by birds, we excluded birds not identified to genus level, those detected flying over the forest and those that did not enter the 30 m radius in point counts. Analyses were conducted at three main levels of response, using generalised linear mixed models (GLMM). The measured habitat variables without the ones excluded due to correlations represented the pool of candidate predictors. This set of plausible predictors was then used to build models using an information-theoretic approach (Burnham and Anderson, 1998), viz., the Akaike Information Criterion (Akaike, 1973, 1974), and more specifically, the corrected AIC (AICC; Hurvich and Tsai, 1989).

First, we looked for patterns of habitat selection at the highest level of all bird species, by summing all the individual species abundances per point count to obtain one abundance value for each count. This ‘bird abundance’ was then used as the response variable in the models. GLMMs were constructed using the lme4 package (Bates et al., 2015), starting with a null model and building up from it. The null model included observer as a fixed effect and sampling points and days as crossed random effects. The residuals were modelled as coming from a Poisson distribution with a log link function. Point ID was used a random effect because of repeated sampling in the same points over the 13 weeks, and sampling day was used as a random effect to account for the random variability in each unique day.

Next, the week variable, which clearly had a strong effect (abundances were higher towards the end of the study period; noticed even during the bird counts), followed by other non-focal (‘nuisance’) variables such as weather, visibility and time of count (on an ordinal scale from 1 to 10 representing which part of the first five hours of sunrise the count was done in) were added to the null models and comparisons were made of AICC values of all the models using the AICctab() function from the bbmle package (Bolker et al., 2020). If the addition of a variable did not result in a substantial decrease of AICC respective to the number of degrees of freedom it used (following the rule of thumb of 2 dAICC per df), it was discarded, and otherwise it was retained in the model. Finally, focal variables were tested including habitat class, presence of moss, log-transformed variables of canopy heterogeneity and tree density and square root-transformed understorey density (transformations done to prevent outliers from gaining too much leverage), scaled proportion of deciduous trees, tree species diversity and richness, and dominant herb. Understorey richness and diversity were not tested because the sampling points did not have sufficient diversity in understorey species and moreover, it was unlikely that birds would track, for instance, spruce saplings (which indeed belonged to my ‘Understorey’ classification). Instead, understorey density was thought to be a meaningful and more important predictor. For continuous or ordinal variables, both linear and polynomial terms were tested since there was no reason to assume a linear relationship with the predictors. If a variable was found to be important, its interaction with ‘Week’ was next tested (i.e., whether or not certain habitat selection criteria changed with the progress of the season).

Such a stepwise procedure was repeated until the final model was obtained. Model diagnostics of all models were explored using the DHARMa package (Hartig, 2020) for distributional assumptions, heteroscedasticity, over-/underdispersion and zero-inflation. Since the model with the default optimiser was facing convergence problems, the BOBYQA optimiser was used which resolved the issue. Once the selection was complete, parsimony was further tested by dropping some complex interaction terms. Finally, the most parsimonious model was obtained.

The next level of response investigated was the guild level. The original bird observations were summarised by guild so as to obtain for each point abundance values of all the guilds that were observed. Only two guilds had high number of overall detections, insectivores (rather, invertebrate-feeders) and omnivores. Models were built separately for these two guilds. Using the lme4 package resulted in several errors and warnings regarding convergence and singular fits, but the same model specifications using the glmmTMB package (Brooks et al., 2017) did not, and hence the latter was used in further steps. The data summarised at guild level was insufficient for the proper utilisation of two crossed random effects (as was done in the first model), so the random effect of day, which had less estimated variance than that of the point, was excluded from the models. This did not result in major differences in the fixed effects estimates so was not considered to be an issue. Model building and checking of model diagnostics was done in the same way as in the first case.

Thirdly, to explore whether patterns differed between migrant and non-migrant species, the bird data was summarised by migration strategy and attempts were made to build GLMMs using the glmmTMB package. However, while a good fitting model was obtained for non-migrants (which had the highest number of overall detections), data for the other two categories seemed to be unsuitable and insufficient. The data showed significant one-inflation when tested (using functions in the DHARMa package) relative to both Poisson and negative binomial distributions, i.e., the number of ‘1’ count observations was too high for a standard Poisson or a standard negative binomial. This one-inflation in turn also caused underdispersion, which was removed neither on switching to distributions suited to underdispersed data (such as the Generalised Poisson and Conway-Maxwell Poisson), nor on introducing a zero-inflation term. The same problem occurred during attempts to build models at the species-level, i.e., looking at patterns of habitat selection for certain selected species (incidentally, the ones with most overall detections). The results of these questions were expected to be interesting, but as the data didn’t seem much suited to a binomial distribution either, this approach was abandoned for lack of better data. Finally, separate models were built to test whether bird abundances were linked to the phenology of the fruiting plants Rubus sp., Vaccinium sp., F. alnus and S. aucuparia but there was no detectable relationship.

All analyses were done using R version ***4.0.2 (R Core Team, 2020)***. Individual figures with multiple graphs were composed using the *patchwork* package for R (Pedersen, 2020).

@@

There were 494 records of birds foraging, although just 2 of them were of birds feeding on fruits. Even in the full unrestricted dataset, there are records of just 14 birds feeding on fruits, and half of these birds were from a single observation. The fruiting plant species most birds fed on was rowan *S. aucuparia* although one record was of a blackbird *Turdus merula* feeding on *Rubus* sp. Most of the birds feeding on fruits were thrushes, but one instance was of a Eurasian blackcap *Sylvia atricapilla* and another of two Eurasian jays *Garrulus glandiarus*. Although this data is very limited and cannot be used in further analyses, the tiny proportion of frugivory among all foraging observations is noteworthy.
